Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-09 infercnv/reports/gains_and_losses.Rmd
fix infercnv.nf bug - used single quotes instead of doubles so variables were not taken; convert Zhang object to sce

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = params$cache_dir)

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)

# dev.off() if there is a device
dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}
```

@@

First, we load the CITE-seq object and extract the metadata and UMAP embeddings.

```{r load_obj, results = 'asis'}
# infercnv output dir
infercnv_dir <- 'out/Zhang2023/by_celltype/'

# load object
obj <- 
  readRDS('/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/all_cells_reference.rds')
```

@@

  # get umap embeddings
  umap <- 
    obj$umap$embedding %>%
    tibble::as_tibble() %>%
    dplyr::bind_cols(
      # add metadata
      obj$meta_data %>%
        tibble::as_tibble() %>%
        dplyr::mutate(id = sample))

@@

  # combine umap, names, and cnvs
  cnvs <-
    umap %>%
    # add cnv metadata
    dplyr::inner_join(cnvs, by = c('id', 'cell')) %>%
    # add names
    dplyr::left_join(names, by = 'name')

@@

```{r}
# function: plot gains and losses
plot_gains_and_losses <- 
  function(cnvs, annot) {
    cnvs %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'has', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(n_cells = dplyr::n(),
                       n_events = sum(value),
                       prop_of_cells = n_events / n_cells) %>%
      ggplot(aes(x = id, y = chr, alpha = prop_of_cells, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines")) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(event ~ annotation)
  }

plot_gains_and_losses(cnvs, 'cell_type')
plot_gains_and_losses(cnvs, 'cluster_name')

# plot celltype
cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
  geom_point(size = 0.1) +
  theme_classic() +
  coord_fixed()

# plot subclusters on umap
cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = subcluster)) +
  geom_point(size = 0.1) +
  theme_classic() +
  coord_fixed()

# plot chr7 gain on umap
cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
  geom_point(size = 0.1) +
  theme_classic() +
  scale_colour_viridis() +
  coord_fixed()
```

@@

p_dat %>%
  ggplot(aes(x = value)) + 
  geom_density() +
  geom_vline(data = p_dat, aes(xintercept = median_proportion)) +
  ggh4x::facet_nested(~ subcluster + cell_type)
```

@@

We pseudobulk chr7+ cells and perform differential expression analysis.

```{r pseudobulk}

```

>>>>

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)
library(Matrix)
library(SummarizedExperiment)
library(SingleCellExperiment)

# dev.off() if there is a device
dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = params$cache_dir) 
```

@@

First, we load the CITE-seq object. We convert it to a SingleCellExperiment.

```{r load_obj, results = 'asis'}
# dirs
data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
infercnv_dir <- 'out/Zhang2023/by_celltype/'

# convert to sce
sce <- xfun::cache_rds({

  # load object
  obj <- 
    readRDS(file.path(data_dir, 'all_cells_reference.rds'))

  # load rna data
  assays <-
    list(
      counts = 
        readRDS(file.path(data_dir, 'raw_mRNA_count_matrix.rds')),
      logcounts = 
        readRDS(file.path(data_dir, 'qc_mRNA_314011cells_log_normalized_matrix.rds')))

  # get cell order from counts
  cell_order <- colnames(assays$logcounts)

  # filter and reorder to cells that pass qc from counts
  assays$counts <- 
    assays$counts[rownames(assays$logcounts), 
                  cell_order]

  # generate coldata
  colData <-
    obj$meta_data[cell_order, ]

  # populate sce
  sce <- 
    SingleCellExperiment::SingleCellExperiment(
      assays = assays,
      colData = colData
    )

  # name, reorder, and add umap embeddings
  umap <- obj$umap$embedding
  rownames(umap) <- rownames(obj$meta_data)
  umap <- umap[cell_order, ]
  reducedDim(sce, 'UMAP_uwot') <- umap

  # return
  sce

}, file = 'sce.rds')

sce <- xfun::cache_rds({

  # get cnv metadata, reorder, left join to colData
  cnvs <- 
    list.files(
      infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(infercnv_dir, file) %>% 
        read.table(sep = '\t')
      }) %>%
    dplyr::bind_rows()

  # left join cnvs to coldata, reorder, restore rownames
  colData <- merge(colData, cnvs, by = 'row.names', all.x = T)
  rownames(colData) <- colData[, 1]
  colData <- colData[, -1]
  colData <- colData[colnames(sce), ]

  # add to sce
  sce@colData <- colData

  # return
  sce

}, file = 'sce_w_infercnv.rds')
```

@@

@@

  # combine umap, names, and cnvs
  cnvs <-
    cnvs %>%
    # add names
    dplyr::left_join(names, by = 'name')

@@

```{r plot_gains_and_losses, fig.height = 20}
# function: plot gains and losses
plot_gains_and_losses <- 
  function(cnvs, annot) {
    cnvs %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'proportion', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(proportion = mean(value)) %>%
      ggplot(aes(x = chr, y = id, alpha = proportion, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines"),
            axis.text.y = element_text(size = 3),
            axis.text.x = element_text(angle = 270, hjust = 0, vjust = 0.5)) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(annotation ~ event)
  }

plot_gains_and_losses(cnvs, 'cell_type')
plot_gains_and_losses(cnvs, 'cluster_name')
```

```{r plot_umaps, fig.height = 15}
list(
  # plot celltype
  cnvs %>%
    dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
    ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
    geom_point(size = 0.05) +
    theme_classic() +
    coord_fixed(),
  # plot chr7 gain on umap
  cnvs %>%
    dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
    dplyr::arrange(value) %>%
    ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
    geom_point(size = 0.05) +
    theme_classic() +
    scale_colour_viridis_c() +
    coord_fixed() +
    ggtitle('proportion of genes duplicated on chr7')
) %>%
  patchwork::wrap_plots(ncol = 1)
```

@@

p_dat %>%
  ggplot(aes(x = value)) + 
  geom_density() +
  geom_vline(data = p_dat, aes(xintercept = median_proportion)) +
  ggh4x::facet_nested(~ subcluster + cell_type)

plot_prop_chr7_gain <- function(p_dat, position_arg) {
  p_dat %>%
    ggplot() +
    geom_bar(aes(x = cell_type, fill = value, group = value),
             position = position_arg) +
    theme_classic() +
    scale_fill_viridis_b(n.breaks = 9) 
}

plot_prop_chr7_gain(p_dat, position_stack(reverse = T))
plot_prop_chr7_gain(p_dat, position_fill(reverse = T))
p_dat %>%
  {split(., .$id)} %>%
  purrr::map2(names(.), ., function(id, df) {
    plot_prop_chr7_gain(df, position_fill(reverse = T)) +
      ggtitle(paste0('chr7 gain count - ', id))
    plot_prop_chr7_gain(df, position_stack(reverse = T)) +
      ggtitle(paste0('chr7 gain proportion - ', id))
  })
```

@@

We pseudobulk chr7+ cells and perform differential expression analysis. In order to do this, we use the `glmGamPoi` package in R. This analysis follows [this tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/glmGamPoi/inst/doc/pseudobulk.html).

```{r pseudobulk}
pseudobulk <- xfun::cache_rds({

  # pseudobulk the sce object by chr7 status
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce, 
      group_by = vars()
    )

}, file = 'pseudobulk.rds')
```
